---
title: "R-Cheat-Sheet"
author: "Clay Rowland"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  
  md_document:
    variant: markdown
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This cheat sheet was developed for a course project in MGSC 790 Data Resource Management, a Business Analytics elective in the [PMBA](https://sc.edu/study/colleges_schools/moore/academic_programs/mba_programs/professional_mba/) program at the University of South Carolina [Darla Moore School of Business](https://sc.edu/study/colleges_schools/moore/index.php).  

It is intended as a quick reference of data structures, functions, and packages for the R language.  Please see the official [Rdocumentation](https://www.rdocumentation.org/) for a comprehensive reference.  

<br />

#### Table of Contents
###### A little bit of Base R

&nbsp;&nbsp;[Data structures](#dataStructures)

&nbsp;&nbsp;[Base functions](#functions)

###### A handful of R packages

&nbsp;&nbsp;[dplyr](#dplyr)

&nbsp;&nbsp;[ggplot2](#ggplot2)

&nbsp;&nbsp;[utils](#utils)

&nbsp;&nbsp;[readr](#readr)

&nbsp;&nbsp;[readxl](#readxl)

&nbsp;&nbsp;[DBI](#dbi)

&nbsp;&nbsp;[httr](#httr)

&nbsp;&nbsp;[jsonlite](#jsonlite)

&nbsp;&nbsp;[haven](#haven)

&nbsp;&nbsp;[RMarkdown](#rMarkdown)

<br />

#### A little bit of Base R

##### <a name="dataStructures"></a>Data structures

A [vector](http://adv-r.had.co.nz/Data-structures.html#vectors) is an array of homogenous datatypes.

A list similar to a vector but can contain heterogeneous datatypes.

A [matrix](http://adv-r.had.co.nz/Data-structures.html#matrices-and-arrays) is a two-dimensional vector of homogenous datatypes.

An array is a vector with one or more dimensions. So, an array with one dimension is (almost) the same as a vector. An array with two dimensions is (almost) the same as a matrix. An array with three or more dimensions is an n-dimensional array.

A [data frame](http://adv-r.had.co.nz/Data-structures.html#data-frames) is similar to a table in a database. Each column in the data frame holds the same type, and the columns can have names.

##### <a name="functions"></a>Base Functions


[c](https://www.rdocumentation.org/packages/base/topics/c)

```{r}
# combine Values into a Vector or List
c("China", "India", "United States of America")

```

[matrix](https://www.rdocumentation.org/packages/base/topics/matrix)
```{r}
# creates a matrix from the given set of values.
matrix(1:9, nrow=3)
```

[data.frame](https://www.rdocumentation.org/packages/base/topics/data.frame)
```{r echo=FALSE}
country = c("China", "India", "United States of America")
population = c(1409517397, 1339180127, 324459463)
```

```{r}
# creates data frames
worldPop <- data.frame(country, population)
worldPop
```

[subset](https://www.rdocumentation.org/packages/base/topics/subset)

```{r}
# return subsets of vectors, matrices or data frames which meet conditions.
subset(worldPop, country=="China")
```

[factor](https://www.rdocumentation.org/packages/base/topics/factor)
```{r}
# encode a vector as a factor
factor(c("A", "B", "B",  "C", "C", "C"))
```

[gl](https://www.rdocumentation.org/packages/base/topics/gl)
```{r}
# generate factor levels
gl(3,1)

```

[levels](https://www.rdocumentation.org/packages/base/topics/levels) 
```{r}
# provides access to the levels attribute of a variable
lowMedHigh <- gl(3, 1)
levels(lowMedHigh) <- c("low", "medium", "high")
lowMedHigh
```


[paste](https://www.rdocumentation.org/packages/base/topics/paste)
```{r}
# concatenate vectors
paste("this", "is", "a", "concatenation")
```

[lapply](https://www.rdocumentation.org/packages/base/topics/lapply) 
```{r}
# apply a function over a list or vector and return a list
lapply(c("a", "b", "c"), toupper)
```

[sapply](https://www.rdocumentation.org/packages/base/topics/sapply) 
```{r}
# apply a function over a list or vector and return a vector, matrix, or array
sapply(c("a", "b", "c"), toupper)
```

[vapply](https://www.rdocumentation.org/packages/base/topics/vapply) 
```{r}
# apply a function over a list or vector and return a vector, matrix, or array of the specified type
vapply(c("a", "b", "c"), toupper, character(1))
```

[identical](https://www.rdocumentation.org/packages/base/topics/identical)
```{r}
x <- 1000
y <- 1000

# evalute whether objects are equal
identical(x, y)
```

[Sys.Date](https://www.rdocumentation.org/packages/base/topics/Sys.Date)
```{r}
# return the current date
Sys.Date
```

[Sys.time](https://www.rdocumentation.org/packages/base/topics/Sys.time)
```{r}
# return the current date and time
Sys.time
```

[format](https://www.rdocumentation.org/packages/base/topics/format)
```{r}
# format an object for printing
format(13.7, nsmall = 3)
```

[summary](https://www.rdocumentation.org/packages/base/topics/summary)
```{r}
# produce result summaries of various model fitting functions
summary(mtcars)
```


<br />

#### A handful of R packages
There are over 13,000 packages that have been developed to supplement the R language.  Below are a handful that were covered in the the course work of the [Data Scientist With R track](https://www.datacamp.com/tracks/data-scientist-with-r) on [Data Camp](https://datacamp.com).  A number of handy packages are provided by the team at [RStudio](https://www.rstudio.com/products/rpackages/).

<br />

<a name="dplyr"></a>[dplyr](http://dplyr.tidyverse.org/)
```{r}
library(dplyr, warn.conflicts = FALSE)

```
dplyr provides a library for data manipulation, including the pipe operator `%>%` which allows for the chaining of functions

```{r}
# filtering and ordering
starwars %>%
  filter(homeworld=="Naboo") %>%
  mutate(taxa = ifelse(species == "Droid", "Machine", "Biological")) %>%
  select(taxa, species, name) %>%
  na.omit() %>%
  arrange(taxa, species, name)

# aggregation
starwars %>%
  filter(homeworld=="Naboo") %>%
  group_by(species) %>%
  tally 
  
```


##### <a name="ggplot2"></a>[ggplot2](http://ggplot2.tidyverse.org/)
```{r}
library(ggplot2)
library(gapminder) # load gapminder dataset
```
ggplot2 provides a library for visualizing data
```{r}
# scatter plot
gapminder %>% filter(year==2007) %>%
ggplot(aes(x=gdpPercap, y=lifeExp, color=continent, size=pop)) +
  geom_point() +  #======  scatter plot
  scale_x_log10() + #=======  logarithmic scale
  theme_minimal() +
  ggtitle("2007 Life Expectancy by GDP Per Capita")

# bar plot
gapminder %>% filter(year==2007) %>%
ggplot(aes(x=continent, y=gdpPercap)) +
  geom_bar(stat="identity", fill="steelblue") + #=====  bar graph
  expand_limits(y=0) +
  theme_minimal() +
  ggtitle("2007 GDP Per Capita by Continent")

# line plot
gapminder %>% filter(country=="China") %>%
ggplot(aes(x=year, y=gdpPercap, color=country)) +
  geom_line(stat="identity", color="blue") + #=====  bar graph
  expand_limits(y=0) +
  theme_minimal() +
  ggtitle("GDP Per Capita by Year in China")


```

##### <a name="httr"></a>[httr](https://www.rdocumentation.org/packages/gdata)
httr provides a library for working with HTTP resources
```{r}
library(httr)
response <- GET("http://httpbin.org/uuid")
status_code(response)
content(response, "text")

```

##### <a name="jsonlite"></a>[jsonlite](https://www.rdocumentation.org/packages/jsonlite)
jsonlite provides a library for parsing and generating JSON
```{r}
library(jsonlite)
fromJSON("http://httpbin.org/json")

gapminder %>% filter(year==2007, country=="China") %>%
  toJSON()

```

##### <a name="gdata"></a>[gdata](https://www.rdocumentation.org/packages/gdata)
gdata provides a library for manipulating data

##### <a name="readr"></a>[readr](https://readr.tidyverse.org/)

readr provides a library for importing tabular data

##### <a name="readxl"></a>[readxl](https://readxl.tidyverse.org/)

readxl provides a library for reading Excel spreadsheets

##### <a name="dbi"></a>[DBI](https://db.rstudio.com/dbi/)

DBI provides a library for connecting to and executing SQL statements against a database

##### <a name="haven"></a>[haven](https://www.rdocumentation.org/packages/haven)
haven provides a library for working with datasets from other statistical software packages (SAS, SPSS, Stata)

##### <a name="rMarkdown"></a> [R Markdown](https://rmarkdown.rstudio.com/)
R Markdown provides a library for generating documents (PDF, HTML, Markdown) that include R functions and output.

This cheat sheet was produced using R Markdown. 

<br />

### References
Data structure and function descriptions have been taken from the [RDocumentation](https://www.rdocumentation.org/) site.

<br />

---
references:
- id: rDocumentation
  title: RDocumentation
  URL: 'http://dx.doi.org/10.1038/nmat3283'
---